HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-06 submission_61.Rmd
Final?

Deciding between man and zone coverage is one of the most critical strategic decisions a defensive coordinator must take before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see a snapshot of the 2024 Week 12 match between the Pittsburgh Steelers and Cleveland Browns). However, in these pre-snap motion does not seem to play an accentuated role (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), although it is a crucial element of modern offensive strategies and it is often used to detect the hidden defensive scheme.

Hence, our contribution explores the potential of the information in pre-snap motion, previously omitted. While we similarly predict man- or zone coverage when the teams are set before snaps, we further leverage the additional information of pre-snap player movements. Specifically, in addition to including rather naive post-motion features, we use a hidden Markov model (HMM) to model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. Incorporating summary statistics of the state decoding results as features into the existing models substantially improves the predictive ability. This lays the groundwork for further analyses such as the evaluation of the effectiveness of pre-snap motion in uncovering defensive strategies.

@@

We aim to forecast the defensive scheme (man- or zone defense) using the `pff_passCoverage` indicator in play-by play data. We omit plays tagged as *others* as well as plays with more than five offensive linemen and with two quarterbacks. Since we are specifically interested in analyzing pre-snap player movements, we concentrate on plays that contain any pre-snap motion. Ultimately, we end up with $3985$ offensive plays in total, from which the defense played $2973$ in zone and $1012$ in man coverage.

@@

To accurately forecast the defensive scheme, we create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: First, using all 11 players on each side, we compute the area spanned by the convex hull of a team as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). Then, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, thus coming up with five players that are a composition of running backs, tight ends and wide receivers. In contrast, for defense, disregarding defense liners (NT, DT, DE), we select the five defenders that were the closest to the five offensive players, corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. From these 10 players, we derive features related to their (standardized) position, distance and orientation.

Additionally, we extract relevant information from the play-by-play data, such as quarter, down, yards to go, home and away score and the remaining game time in the current half (in seconds). See the Appendix for a more detailed description of the features.

@@

In general, we have a limited dataset available (only 3985 plays) and therefore need to manage model complexity by controlling the number of features. Given the small dataset, we focus on 32 previously described basic features used in all 3 models: 6 convex hull related features, 20 player features, and 6 play-by-play features. In the Appendix, we provide a discussion on the choice of features.

First of all, we need a suitable basic model class for predicting man or zone coverage and opt for the following two: First, we fit a `glmnet` (elastic net) model, which performs implicit feature selection and is able to handle multicollinearity. Second, we use an `xgboost` model, which is able to capture non-linear effects (and interactions) and also handles multicollinearity. However, it necessitates careful hyperparameter tuning and generally performs better on bigger data sets. For all of the models, we use 10-fold cross validation on a suitable hyperparameter grid.

@@

At first, we fit the aforementioned models with the previously described basic pre-motion features as all of these features are derived at the time of line-set. These very basic models serve as baseline models that allows to measure the effect of pre-snap motion (features) in the following.

@@

In a second step, we extend our basic pre-motion model with naive post-motion features. To keep the complexity manageable, we derive only six additional post-motion features: for each team (offense and defense), we infer the maximum $y$-distance, the maximum $x$-distance and the total distance traveled by both teams until the snap.

@@

The previously described models primarily serve as baselines for evaluating the impact of incorporating pre-snap motion information into predicting defensive schemes. The main contribution of our project lies in the effective integration of this information using a hidden Markov model (HMM).

Specifically, we use results derived from an HMM fitted to the pre-snap motion tracking data as further features in the previously described models. To achieve this, we model the movements of the five defensive players during pre-snap motion by an HMM (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a Gaussian distribution with mean aligned to the $y$-coordinate of the offensive player they are guarding and an estimated standard deviation. HMMs are particularly well-suited for this task since the guarding assignments are not directly observed but need to be inferred from the defenders' responses to the offenders' movements.
Hence, the model naturally treats this information as a latent state variable, making each observation a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
The ultimate goal is to use the fitted model for *state-decoding*, i.e., inferring information on the guarding assignment based on the observations made. HMMs excel in this context as they leverage not only the defenders' $y$-coordiate and the $y$-coordiate of all offenders potentially guarded at the current time point, but also incorporate probabilistic information based on the previous and subsequent observations. State-decoding can either provide us with fully probabilistic information, i.e., the probability of each defender guarding each offensive player at each time point, or a single most likely guarding assignment for each defender at each time point. We will use the former to enhance our post-motion models and the latter for visualization purposes.
For details on model fitting and state decoding, we also refer to the Appendix.

@@

<center>
  <img src="./animation/animation.gif" width="423" height="450" alt="Coverage Prediction according to a hidden Markov model.">
</center>

@@

To incorporate pre-snap motion information, we re-train the post-motion model to predict whether the defense employs a man- or zone-coverage scheme, now integrating results from the HMM analysis as additional features. A caveat of the aforementioned decoded state probabilities is that they form a multi-dimensional time series, complicating their direct inclusion as features in the previously described models. This challenge can however be remedied by employing suitable summary statistics. To achieve this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$, where $T_n$ is the length of the $n$-th play, and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $t$ and $t+1$. From this, we can calculate some simple summary statistics, specifically 1) the sum of state switches, 2) the average number of state switches, and 3) the number of defenders that switch offensive players during a play. Additionally, we use the aforementioned decoded state probabilities to calculate a more elaborated statistic, the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$:

@@

As the entropy is a measure of uncertainty or randomness in a probability distribution, higher entropy indicates greater unpredictability, while lower entropy signifies more predictability. In our setting, we suspect that higher entropy values are associated with less persistent guarding allocations, while lower entropy values are indicative of more stable guarding assignments.

@@

`r colorize("Den Text finde ich irgendwie unrund und nicht so leicht verständlich.","red")`

<!--
We compare the predictive performance of our three models to determine the effectiveness of player movements before snap for detecting the correct defensive scheme. We train the models on 85% of the data and evaluate performance on the remaining 15%. Since we use cross-validation for tuning of the models during the training phase, we perform our evaluation repeatedly. We use 50 repetition each time splitting the data differently, and for each we tune and train a model and then evaluate the performance on the test set. We do this to ensure that our results are meaningful and not due to chance (especially in our small data set).
-->

`r colorize("Vorschlag (Robert):","red")`

To assess whether the incorporation of pre-snap player movements can improve the prediction of defensive schemes, we evaluate the predictive performance of our three models. Therefore, we perform a train-test split using 85% for training and 15% for testing. To ensure robust and reliable results, we repeat the evaluation process 50 times, using different CV splits for each iteration. For each split, we train and fine-tune the model on the training data and then evaluate its performance on the test set. This approach helps minimize the risk of our findings being influenced by randomness, which is particularly important given the limited size of our dataset. The results of the evaluation are shown in the figures below.

<!-- We evaluate the predictive performance of our three models to assess how well pre-snap player movements help identify the correct defensive scheme. To do this, we split the data into two sets: 85% for training and 15% for testing. During the training phase, we use cross-validation to fine-tune the models. -->

<!-- To ensure robust and reliable results, we repeat the evaluation process 50 times, using different cv splits for each iteration. For each split, we train and fine-tune the model on the training data and then evaluate its performance on the test set. This approach helps minimize the risk of our findings being influenced by randomness, which is particularly important given the limited size of our dataset. -->

`r colorize("Die plots sind noch sehr verpixelt","red")`

`r colorize("Robert: komisch, bei mir sind sie eig recht scharf... hab versucht es bisschen zu ändern. Wenns das nicht besser macht, weiß ich leider auch nich was hilft...","red")`

<div style="text-align: center;">
  ![](ACCs_bdb_main.png)
  ![](AUCs_bdb_main.png)
</div>

<!-- ![](ACCs_bdb_main.png) -->
<!-- ![](AUCs_bdb_main.png) -->

The two figures above show the results from our 50 repetitions of the experiment. The first figure illustrates the classification accuracy, computed using a threshold of 50\%. Since this value is somewhat arbitrary and our dataset is slightly imbalanced ($\approx$ 75\% zone coverage plays and $\approx$ 25\% man coverage plays), accuracy is a suboptimal metric for evaluating the results. Hence, we also present the AUC values of the three models (see Appendix for a discussion on evaluation metrics) in the second figure. 
Several observations emerge from these metrics. On the one hand, the more flexible `xgboost` model outperforms `glmnet`, but it also exhibits greater variation, indicating that additional data could enhance the tuning of this model. 
On the other hand, the pre-motion model clearly performs worst. Once motion information is incorporated, we observe a substantial improvement in performance.
While this is to be expected, importantly, we also see that the HMM post-motion model outperforms the model only using naive post-motion information. This indicates that the features derived from the HMM's state decoding capture distinct information beyond simple motion features, providing an additional dimension that enhances the overall analysis.

@@

`r colorize("Die figure fehlt irgendwie noch","red")`

![](differences_teams.png){width=50%}

Main findings:

Plot: In general, motion improves the correct detection of coverage (median of most teams positive). Table: Giants (Titans) --> low number of motions but very effective 49 out of 60 (82 %) of the motion plays increased coverage detections. Miami, Atlanta, SF --> Extensive use of motion but still very effective. Raider, Seattle, Rams --> Substantial use of motion but very ineffective. Bills (Bengals) -->  little use of motion and also ineffective.
`r colorize("Rouven, wenn Du es schaffst einen netten text zu den Findings zu verfassen wäre es super!","red")`

@@

This project explored the use of hidden Markov models (HMMs) to enhance the prediction of defensive schemes in football by incorporating pre-snap player movements. Models augmented with HMM-derived features demonstrated improved predictive performance compared to those relying solely on naive motion data.

Comparing predictions from pre-motion and post-motion models provides insights into an offense's ability to use motion effectively to identify defensive coverage schemes. Similarly, future analysis could also relate to evaluating a defense's skill in disguising their coverage and confusing the offense during pre-snap motion.

>>>>

Deciding between man and zone coverage is one of the most critical strategic decisions a defensive coordinator must take before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see snapshot below). However, pre-snap motion does not seem to play an accentuated role in this model (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), although it is a crucial element of modern offensive strategies.

Hence, our contribution explores the potential of incorporating pre-snap motion. While we similarly predict man- or zone coverage before motion, we further leverage the additional information of pre-snap player movements. Specifically, in addition to including rather naive post-motion features, we use an HMM to model defenders' trajectories based on hidden states, which represent the offensive players they may be guarding. Incorporating summary statistics of the state decoding results as features into the existing models substantially improves the predictive ability. This lays the groundwork for further analyses such as the evaluation of the effectiveness of pre-snap motion in uncovering defensive strategies.

@@

We aim to forecast the man- or zone coverage using the `pff_passCoverage` indicator in play-by play data. We omit plays tagged as *others* and plays with more than five offensive linemen and with two quarterbacks. Since we are specifically interested in analyzing pre-snap player movements, we concentrate on plays that contain any pre-snap motion. Ultimately, we end up with $3985$ offensive plays in total, from which the defense played $2973$ in zone and $1012$ in man coverage.

@@

To accurately forecast the defensive scheme, we create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: First, using all 11 players on each side, we compute the area spanned by the convex hull of a team and the largest $y$ distance (i.e., the width of the hull) and the largest $x$ distance (i.e., the length of the hull), i.e. 6 features. Then, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB and for defense, we disregard defense lineman (NT, DT, DE) and select the five defenders that were the closest to the five offensive players, corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. From these 10 players, we derive 20 features related to their (standardized) position.

Additionally, we extract information from the play-by-play data, such as quarter, down, yards to go, home and away score and the remaining seconds in the current half. See the Appendix for a more detailed description and a discussion on the choice of features.

@@

In general, we have a limited dataset available (only 3985 plays) and therefore need to manage model complexity by controlling the number of features. Given the small dataset, we focus on the 32 previously described basic features.

Regarding a suitable basic model class for predicting man or zone coverage, we opt for the following two: First, we fit a `glmnet` (elastic net) model, which performs implicit feature selection and can handle multicollinearity. Second, we use an `xgboost` model, which additionally captures non-linear effects (and interactions). For these models, we use 10-fold cross validation on a suitable hyperparameter grid.

@@

First, we fit the aforementioned models with the previously described basic pre-motion features. These very basic models serve as baseline models that allows to measure the effect of pre-snap motion (features) in the following.

@@

Second, we extend our basic pre-motion model with naive post-motion features. To keep the complexity manageable, we derive only six additional post-motion features: for each team (offense and defense), we infer the maximum $y$-distance, the maximum $x$-distance and the total distance traveled by both teams until the snap.

@@

The previously described models primarily serve as baselines for evaluating the impact of incorporating pre-snap motion information into predicting defensive schemes. The main contribution of our project lies in the effective integration of this information using an HMM.

Specifically, we use results derived from an HMM fitted to the pre-snap motion tracking data as additional features in the previously described models. To achieve this, we model the movements of the five defensive players during pre-snap motion by an HMM (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a Gaussian distribution with mean aligned to the $y$-coordinate of the offensive player they are guarding and an estimated standard deviation. HMMs are particularly well-suited for this task since the guarding assignments are not directly observed but need to be inferred from the defenders' responses to the offenders' movements.
Hence, the model naturally treats this information as a latent state variable, making each observation a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
The ultimate goal is to use the fitted model for *state-decoding*, i.e., inferring information on the guarding assignment based on the observations made. HMMs excel in this context as they leverage not only the defenders' $y$-coordiate and the $y$-coordiate of all offenders potentially guarded at the current time point, but also incorporate probabilistic information based on the previous and subsequent observations. State-decoding can either provide us with fully probabilistic information, i.e., the probability of each defender guarding each offensive player at each time point, or a single most likely guarding assignment for each defender at each time point. We will use the former to enhance our post-motion models and the latter for visualization purposes.
For details on model fitting and state decoding, we also refer to the Appendix.

@@

<center>
  <img src="./animation/animation.gif" width="423" height="450" alt="Coverage Prediction according to the HMM.">
</center>

@@

Third, we re-train the models from above, now integrating HMM results as additional features. A caveat of the aforementioned decoded state probabilities is that they form a multi-dimensional time series, complicating their direct inclusion as features in the previously described models. This challenge can however be remedied by employing suitable summary statistics. To achieve this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$, where $T_n$ is the length of the $n$-th play, and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $t$ and $t+1$. From this, we can calculate some simple summary statistics, specifically 1) the sum of state switches, 2) the average number of state switches, and 3) the number of defenders that switch offensive players during a play. Additionally, we use the aforementioned decoded state probabilities to calculate a more elaborated statistic, the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$:

@@

As the entropy is a measure of uncertainty in a probability distribution, lower (higher) entropy indicates greater (un)predictability. Here, we suspect that higher (lower) entropy values are associated with less (more stable) persistent guarding allocations.

@@

To assess whether the incorporation of pre-snap player movements can improve the prediction of defensive schemes, we evaluate the predictive performance of our three models using a train-test split of 85/15. To ensure robust and reliable results, we repeat the evaluation process 50 times, using different CV splits for each iteration. For each split, we train and fine-tune the model on the training data and then evaluate its performance on the test set. This approach helps minimize the risk of our findings being influenced by randomness, which is particularly important given the limited size of our dataset. The results of the evaluation are shown in the figures below.

<div style="text-align: center;">
  ![](ACCs_bdb_main2.png){width=50%}
  ![](AUCs_bdb_main2.png){width=50%}
</div>

The first figure illustrates the classification accuracy, computed using a threshold of 50\%. Since this value is somewhat arbitrary and our dataset is slightly imbalanced ($\approx$ 75\% zone and $\approx$ 25\% man coverage plays), accuracy is a suboptimal metric (see Appendix for a discussion on evaluation metrics). Hence, we also present the AUC values of the three models in the second figure. Several observations emerge from these metrics. First, the more flexible `xgboost` model outperforms `glmnet`, but also exhibits greater variation, indicating that additional data could enhance the tuning of this model. 
Second, the pre-motion model clearly performs worst. Once motion information are incorporated, we observe a substantial improvement in performance.
While this is to be expected, importantly, we also see that the HMM post-motion model outperforms the model only using naive post-motion information. This indicates that the features derived from the HMM's state decoding capture distinct information beyond simple motion features, providing an additional dimension that enhances the overall analysis.

@@

<!-- ![](differences_teams.png){width=50%} -->

Our approach is not limited to the mere prediction of coverages; it can also be applied to a variety of analyses. As an illustrative example, we focuse on examining how pre-snap motion has helped teams, according to our model, identify the correct defensive coverage. Specifically, we calculate pre- and HMM post-motion probabilities for each team and evaluated the extent to which these models accurately predicted the actual coverages.

As shown in the following table, teams that are well-known for extensive use of pre-snap motion --- such as the Niners and Dolphins --- are also among the ones that most frequently increased the likelihood of correctly decoding the defensive coverage. While some teams, such as the Giants or Titans, succeed in accurately identifying coverages despite employing limited motion, the general trend indicates that teams utilizing extensive motion are more effective at leveraging it to recognize defensive schemes.

<!-- Plot: In general, motion improves the correct detection of coverage (median of most teams positive). Table: Giants (Titans) --> 
<!-- low number of motions but very effective 49 out of 60 (82 %) of the motion plays increased coverage detections. Miami, Atlanta, SF --> 
<!-- Extensive use of motion but still very effective. Raider, Seattle, Rams --> 
<!-- Substantial use of motion but very ineffective. Bills (Bengals) -->  
<!-- little use of motion and also ineffective. -->
<!-- `r colorize("Rouven, wenn Du es schaffst einen netten text zu den Findings zu verfassen wäre es super!","red")` -->

@@

This project explored the use of HMMs to enhance the prediction of defensive schemes by incorporating pre-snap player movements. Models augmented with HMM-derived features demonstrated improved predictive performance compared to those relying solely on naive motion data.
